---
output: html_document
editor_options: 
  chunk_output_type: console
---
Analysis of pilot data
analysisID: agan0005

Define the names and location of input data
•If you change the names of these files, you can run a different dataset through without needed to change the rest of the code•

```{r}

#setwd(dir = "/Users/alexgibbs/OneDrive - Cardiff University/crest/analyses/agan0005/bin") #set your working directory

metaboliteFile <- "metabolte.csv"
speciesFile <- "microbe_species.csv"
proteinFile <- "protein_ID.csv"
manifestFile <- "metadata.2groups.xlsx"

metaboliteFile <- paste0("../input/", metaboliteFile)
speciesFile <- paste0("../input/", speciesFile)
proteinFile <- paste0("../input/", proteinFile)
manifestFile <- paste0("../input/", manifestFile)

```

install required packages

```{r}

if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")
  BiocManager::install("mixOmics")

## reading excel files

install.packages("readxl")

# data wrangling

install.packages("dplyr")

# plotting

install.packages("ggplot2")
install.packages("ggridges")

```

Load library to begin analysis

```{r}

library("mixOmics")
library("readxl")
library("dplyr")
library("ggplot2")
library("ggridges")

```


Read in metabolite data

```{r}

## metabolite IDs are name the rownames of the dataframe
metaboliteData <- read.csv(metaboliteFile, row.names = 1, header = TRUE, check.names = F)

## remove the first row of sampleTypes
metaboliteData <- metaboliteData[-1, ]
rNames <- row.names(metaboliteData)

## change character to numeric 
metaboliteData <- as.data.frame(sapply(metaboliteData, as.numeric))
row.names(metaboliteData) <- rNames

## check the dimension of the dataframe as a check
print(dim(metaboliteData))

```

Read in protein data

```{r}

## metabolite IDs are name the rownames of the dataframe
proteinData <- read.csv(proteinFile, row.names = 1, header = TRUE, check.names = F)

## remove first column of protein names
proteinData[,1] <- NULL

## remove the first row of sampleTypes
proteinData <- proteinData[-1, ]
rNames <- row.names(proteinData)

## change character to numeric 
proteinData <- as.data.frame(sapply(proteinData, as.numeric))
row.names(proteinData) <- rNames

## check the dimension of the dataframe
print(dim(proteinData))

```

Read in species data

```{r}

## metabolite IDs are name the rownames of the dataframe
speciesData <- read.csv(speciesFile, row.names = 1, header = TRUE, check.names = F)

## remove the first row of sampleTypes
speciesData  <- speciesData[-1, ]
rNames <- row.names(speciesData)

## change character to numeric 
speciesData <- as.data.frame(sapply(speciesData, as.numeric))
row.names(speciesData) <- rNames

## check the dimension of the dataframe
print(dim(speciesData))

```

Read in sample manifest 

```{r}

sampleKey <- read_excel(manifestFile)

#change the column names

colnames(sampleKey) <- c("sampleName", "sampleGroup")

```

Order columns by sampleName for all dataframes, so that samples are in the same order

```{r}

metaboliteData <- metaboliteData[sampleKey$sampleName]
proteinData <- proteinData[sampleKey$sampleName]
speciesData <- speciesData[sampleKey$sampleName]

## check to make sure in correct order
identical(colnames(metaboliteData), sampleKey$sampleName)
identical(colnames(proteinData), sampleKey$sampleName)
identical(colnames(speciesData), sampleKey$sampleName)

```

Create numerical arrays from the dataframes
•These need to be transposed so that samplenames are now rownames• do this later

```{r}

mArray <- as.matrix((metaboliteData))
pArray <- as.matrix((proteinData))
sArray <- as.matrix((speciesData))

```

Write these numerics back to a file under output (although we don't use these)

```{r}

write.table(mArray, file="../output/raw/metabolite.raw.txt", row.names=T, col.names=T, sep="\t")
write.table(pArray, file="../output/raw/protein.raw.txt", row.names=T, col.names=T, sep="\t")
write.table(sArray, file="../output/raw/species.raw.txt", row.names=T, col.names=T, sep="\t")

```

NORMALISATION: species data
Transform each column separately from counts to percentage of total counts

```{r}

# create empty array for percentage data

normSArray <- sArray

# for each column in turn

for (i in 1:length(colnames(sArray))) {
  normSArray[,i] <- sapply(sArray[,i], function(x) (x / sum(sArray[,i])) * 100 )
}

# write to file for checking

write.table(normSArray, file="../output/norm/species.norm.txt", row.names=T, col.names=T, sep="\t")

```

NORMALISATION: metabolite data
Community standard: log10 transform

```{r}

# log10

normMArray <- log(mArray, base=10)

# restore zero values from Inf

normMArray[!is.finite(normMArray)] <- 0

# write to file for checking

write.table(normMArray, file="../output/norm/metabolite.norm.txt", row.names=T, col.names=T, sep="\t")

```

NORMALISATION: protein data
Community standard: log10 transform

PROPOSED CHANGE: https://bioconductor.org/packages/release/bioc/vignettes/DEP/inst/doc/DEP.html#normalization
Notes on proposed change: this package is for the differential expression of protein between groups. it takes the data from quantitative analyses of raw mass spec data as input.
the analysis generates a summarizedExperiment object, filters missing values (proteins not quantified in all replicates). The data is firstly log2 transformed, then variance stabilised (vsn)

OBERVATION: this is zero-centric data
2 populations in the data: a zero population and a not zero population
Therefore the anova is suboptimal, hence need to look at bioconductor package

hist(normPArray)
hist(normPArray[,1])

```{r}

# log10

normPArray <- log(pArray, base=10)

# restore zero values from Inf

normPArray[!is.finite(normPArray)] <- 0

# write to file for checking

write.table(normPArray, file="../output/norm/protein.norm.txt", row.names=T, col.names=T, sep="\t")

```

boxplots

```{r}

## define a function for plotting

myBoxplot <- function(aa, datatype, title, xtitle, ytitle) {
  
  plotMe <- stack(as.data.frame(aa))
  
  ggplot(plotMe) +
    geom_boxplot(aes(x = ind, y = values)) +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
    ggtitle(title) +
    xlab(xtitle) +
    ylab(ytitle)
  
  ggsave(paste0("../output/", datatype, "/", gsub(" ", "_", title), ".boxplot.", datatype, ".pdf"))
                
}

# protein

myBoxplot(pArray, "raw", "raw protein", "sampleID", "raw abundance")
myBoxplot(normPArray, "norm", "log10 normalised protein", "sampleID", "log10 abundance")

# metabolite

myBoxplot(mArray, "raw", "raw metabolite", "sampleID", "raw abundance")
myBoxplot(normMArray, "norm", "log10 normalised metabolite", "sampleID", "log10 abundance")

# species

myBoxplot(sArray, "raw", "raw species", "sampleID", "species count")
myBoxplot(normSArray, "norm", "percentage species count", "sampleID", "percentage count")

```

distribution plots (these take a while to plot)

```{r}

## define a function for plotting

myDisplot <- function(aa, datatype, title, xtitle, ytitle) {
  
  plotMe <- stack(as.data.frame(aa))
  plotMe <- data.frame(ind=plotMe$ind, values=plotMe$values)
  plotMe <- plotMe %>% mutate_all(as.ordered)

  ggplot(plotMe, aes(x = values, y = ind)) +
    geom_density_ridges2() 

  ggsave(paste0("../output/", datatype, "/", gsub(" ", "_", title), ".freqplot.", datatype, ".pdf"))
                
}

# protein

#myDisplot(pArray, "raw", "raw protein", "sampleID", "raw abundance")
#myDisplot(normPArray, "norm", "log10 normalised protein", "sampleID", "log10 abundance")

# metabolite

#myDisplot(mArray, "raw", "raw metabolite", "sampleID", "raw abundance")
#myDisplot(normMArray, "norm", "log10 normalised metabolite", "sampleID", "log10 abundance")

# species

#myDisplot(normSArray, "norm", "percentage species count", "sampleID", "percentage count")
#myDisplot(sArray, "raw", "raw species", "sampleID", "species count")

```

Shapiro-Wilk test for normality
https://www.r-bloggers.com/2019/08/shapiro-wilk-test-for-normality-in-r/

```{r}

# metabolite

shapiroWilkPvalues <- data.frame(rawMetabolite=apply(mArray, 2, function(x) shapiro.test(x)$p.value))

shapiroWilkPvalues$normMetabolite <- apply(normMArray, 2, function(x) shapiro.test(x)$p.value)
        
# protein

shapiroWilkPvalues$rawProtein <- apply(pArray, 2, function(x) shapiro.test(x)$p.value)

shapiroWilkPvalues$normProtein <- apply(normPArray, 2, function(x) shapiro.test(x)$p.value)

# species

shapiroWilkPvalues$rawSpecies <- apply(sArray, 2, function(x) shapiro.test(x)$p.value)

shapiroWilkPvalues$normSpecies <- apply(normSArray, 2, function(x) shapiro.test(x)$p.value)


# write to file

shapiroWilkPvalues <- cbind(row.names(shapiroWilkPvalues), shapiroWilkPvalues)
colnames(shapiroWilkPvalues)[1] <- "sampleID"
write.table(shapiroWilkPvalues, file="../output/shapirowilks.txt", row.names=F, col.names=T, sep="\t")

```

