---
output: html_document
editor_options: 
  chunk_output_type: console
---
Analysis of pilot data
analysisID: agan0005

Define the names and location of input data
•If you change the names of these files, you can run a different dataset through without needed to change the rest of the code•

```{r}

#setwd(dir = "set/your/path") #set your working directory

metaboliteFile <- "metabolte.csv"
speciesFile <- "microbe_species.csv"
proteinFile <- "protein_ID.csv"
manifestFile <- "metadata.2groups.xlsx"

metaboliteFile <- paste0("../input/", metaboliteFile)
speciesFile <- paste0("../input/", speciesFile)
proteinFile <- paste0("../input/", proteinFile)
manifestFile <- paste0("../input/", manifestFile)

```

install required packages

```{r}

if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")
  BiocManager::install("mixOmics")

## reading excel files

install.packages("readxl")

# data wrangling

install.packages("dplyr")

# plotting

install.packages("ggplot2")
install.packages("ggridges")

```

Load library to begin analysis

```{r}

library("mixOmics")
library("readxl")
library("dplyr")
library("ggplot2")
library("ggridges")

```


Read in metabolite data

```{r}

## metabolite IDs are name the rownames of the dataframe
metaboliteData <- read.csv(metaboliteFile, row.names = 1, header = TRUE, check.names = F)

## remove the first row of sampleTypes
metaboliteData <- metaboliteData[-1, ]
rNames <- row.names(metaboliteData)

## change character to numeric 
metaboliteData <- as.data.frame(sapply(metaboliteData, as.numeric))
row.names(metaboliteData) <- rNames

## check the dimension of the dataframe as a check
print(dim(metaboliteData))

```

Read in protein data

```{r}

## metabolite IDs are name the rownames of the dataframe
proteinData <- read.csv(proteinFile, row.names = 1, header = TRUE, check.names = F)

## remove first column of protein names
proteinData[,1] <- NULL

## remove the first row of sampleTypes
proteinData <- proteinData[-1, ]
rNames <- row.names(proteinData)

## change character to numeric 
proteinData <- as.data.frame(sapply(proteinData, as.numeric))
row.names(proteinData) <- rNames

## check the dimension of the dataframe
print(dim(proteinData))

```

Read in species data

```{r}

## metabolite IDs are name the rownames of the dataframe
speciesData <- read.csv(speciesFile, row.names = 1, header = TRUE, check.names = F)

## remove the first row of sampleTypes
speciesData  <- speciesData[-1, ]
rNames <- row.names(speciesData)

## change character to numeric 
speciesData <- as.data.frame(sapply(speciesData, as.numeric))
row.names(speciesData) <- rNames

## check the dimension of the dataframe
print(dim(speciesData))

```

Read in sample manifest 

```{r}

sampleKey <- read_excel(manifestFile)

#change the column names

colnames(sampleKey) <- c("sampleName", "sampleGroup")

```

Order columns by sampleName for all dataframes, so that samples are in the same order

```{r}

metaboliteData <- metaboliteData[sampleKey$sampleName]
proteinData <- proteinData[sampleKey$sampleName]
speciesData <- speciesData[sampleKey$sampleName]

## check to make sure in correct order
identical(colnames(metaboliteData), sampleKey$sampleName)
identical(colnames(proteinData), sampleKey$sampleName)
identical(colnames(speciesData), sampleKey$sampleName)

```

Create numerical arrays from the dataframes
•These need to be transposed so that samplenames are now rownames• do this later

```{r}

mArray <- as.matrix((metaboliteData))
pArray <- as.matrix((proteinData))
sArray <- as.matrix((speciesData))

```

Write these numerics back to a file under output (although we don't use these)

```{r}

write.table(mArray, file="../output/raw/metabolite.raw.txt", row.names=T, col.names=T, sep="\t")
write.table(pArray, file="../output/raw/protein.raw.txt", row.names=T, col.names=T, sep="\t")
write.table(sArray, file="../output/raw/species.raw.txt", row.names=T, col.names=T, sep="\t")

```

NORMALISATION: species data
Transform each column separately from counts to percentage of total counts

```{r}

# create empty array for percentage data

normSArray <- sArray

# for each column in turn

for (i in 1:length(colnames(sArray))) {
  normSArray[,i] <- sapply(sArray[,i], function(x) (x / sum(sArray[,i])) * 100 )
}

# write to file for checking

write.table(normSArray, file="../output/norm/species.norm.txt", row.names=T, col.names=T, sep="\t")

```

NORMALISATION: metabolite data
Community standard: log10 transform

```{r}

# log10

normMArray <- log(mArray, base=10)

# restore zero values from Inf

normMArray[!is.finite(normMArray)] <- 0

# write to file for checking

write.table(normMArray, file="../output/norm/metabolite.norm.txt", row.names=T, col.names=T, sep="\t")

```

NORMALISATION: protein data
Community standard: log10 transform

PROPOSED CHANGE: https://bioconductor.org/packages/release/bioc/vignettes/DEP/inst/doc/DEP.html#normalization

OBERVATION: this is zero-centric data
2 populations in the data: a zero population and a not zero population
Therefore the anova is suboptimal, hence need to look at bioconductor package

hist(normPArray)
hist(normPArray[,1])

```{r}

# log10

normPArray <- log(pArray, base=10)

# restore zero values from Inf

normPArray[!is.finite(normPArray)] <- 0

# write to file for checking

write.table(normPArray, file="../output/norm/protein.norm.txt", row.names=T, col.names=T, sep="\t")

```

boxplots

```{r}

## define a function for plotting

myBoxplot <- function(aa, datatype, title, xtitle, ytitle) {
  
  plotMe <- stack(as.data.frame(aa))
  
  ggplot(plotMe) +
    geom_boxplot(aes(x = ind, y = values)) +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
    ggtitle(title) +
    xlab(xtitle) +
    ylab(ytitle)
  
  ggsave(paste0("../output/", datatype, "/", gsub(" ", "_", title), ".boxplot.", datatype, ".pdf"))
                
}

# protein

myBoxplot(pArray, "raw", "raw protein", "sampleID", "raw abundance")
myBoxplot(normPArray, "norm", "log10 normalised protein", "sampleID", "log10 abundance")

# metabolite

myBoxplot(mArray, "raw", "raw metabolite", "sampleID", "raw abundance")
myBoxplot(normMArray, "norm", "log10 normalised metabolite", "sampleID", "log10 abundance")

# species

myBoxplot(sArray, "raw", "raw species", "sampleID", "species count")
myBoxplot(normSArray, "norm", "percentage species count", "sampleID", "percentage count")

```

distribution plots (these take a while to plot)

```{r}

## define a function for plotting

myDisplot <- function(aa, datatype, title, xtitle, ytitle) {
  
  plotMe <- stack(as.data.frame(aa))
  plotMe <- data.frame(ind=plotMe$ind, values=plotMe$values)
  plotMe <- plotMe %>% mutate_all(as.ordered)

  ggplot(plotMe, aes(x = values, y = ind)) +
    geom_density_ridges2() 

  ggsave(paste0("../output/", datatype, "/", gsub(" ", "_", title), ".freqplot.", datatype, ".pdf"))
                
}

# protein

#myDisplot(pArray, "raw", "raw protein", "sampleID", "raw abundance")
#myDisplot(normPArray, "norm", "log10 normalised protein", "sampleID", "log10 abundance")

# metabolite

#myDisplot(mArray, "raw", "raw metabolite", "sampleID", "raw abundance")
#myDisplot(normMArray, "norm", "log10 normalised metabolite", "sampleID", "log10 abundance")

# species

#myDisplot(normSArray, "norm", "percentage species count", "sampleID", "percentage count")
#myDisplot(sArray, "raw", "raw species", "sampleID", "species count")

```

Shapiro-Wilk test for normality
https://www.r-bloggers.com/2019/08/shapiro-wilk-test-for-normality-in-r/

```{r}

# metabolite

shapiroWilkPvalues <- data.frame(rawMetabolite=apply(mArray, 2, function(x) shapiro.test(x)$p.value))

shapiroWilkPvalues$normMetabolite <- apply(normMArray, 2, function(x) shapiro.test(x)$p.value)
        
# protein

shapiroWilkPvalues$rawProtein <- apply(pArray, 2, function(x) shapiro.test(x)$p.value)

shapiroWilkPvalues$normProtein <- apply(normPArray, 2, function(x) shapiro.test(x)$p.value)

# species

shapiroWilkPvalues$rawSpecies <- apply(sArray, 2, function(x) shapiro.test(x)$p.value)

shapiroWilkPvalues$normSpecies <- apply(normSArray, 2, function(x) shapiro.test(x)$p.value)


# write to file

shapiroWilkPvalues <- cbind(row.names(shapiroWilkPvalues), shapiroWilkPvalues)
colnames(shapiroWilkPvalues)[1] <- "sampleID"
write.table(shapiroWilkPvalues, file="output/shapirowilks.txt", row.names=F, col.names=T, sep="\t")

```


http://mixomics.org/mixdiablo/

integrate numerical data into a list for diablo

```{r}

allData <- list(metabolite=mArray, protein=pArray, species=sArray)

```

run diablo method
we first fit a block PSDA model
```{r}

## require the near.zero.var to take account of protein features with zero variance (a lot of zero values)

diabloResults <- block.plsda(allData, sampleKey$sampleGroup, near.zero.var = T)

```

Plot Diablo
visualise the correlation between the first components from each dataset. numbers indicate the pearsons correlation coefficient. 95% confidence elipse plots.
```{r}

plotDiablo(diabloResults)

```


Plot samples
plots the individual PCA plots.
```{r}

plotIndiv(diabloResults,
          legend = T) 

```

Plot Arrows
start of the arrow indicates the centroind between all datasets for a given sample and the tip of the arrow the location of that same sample but in each block. plot highlights the agreement between all datasets at the sample level
```{r}

plotArrow(diabloResults)


```

Plot variables
Correlation circle plot indicates contribution of each variable to each component. important variables should be close to the larger circle.
clustering of points indicate a strong correlation between variables

as well as correlation between variables

```{r}

## size of point changed with cex

plotVar(diabloResults, var.names=F, style='graphics', legend=T, pch = c(16, 17, 15), cex=c(0.5, 0.5, 0.5), col = c("darkorchid", "brown1", "lightgreen"), title="pilot")

```

Circos plot
represents correlations between variables of different types, represented on the side quadrants
several display options are possible. to show within and between connections between blocks, and expression levels of each variable according to each class, use line=TRUE
cutoff argument is used to visualise correlation coefficients above this threshold in the multi-omics signature

the plot allows us to visualise cross-correlations between data types, and the nature of these correlations (pos or neg)
```{r}

circosPlot(diabloResults, cutoff = 0.9, line = TRUE, color.blocks = c('darkorchid', 'brown1', 'lightgreen'), color.cor = c("grey20", "blue"), size.labels = 1.5)

```

Network
relevance networks,also built on similarity matrix, can also visualise correlations between different types of variables. each colour represents a variable type.
threshold can be set using argument cutoff

```{r}
network(diabloResults,
        cutoff = 0.5)



```



