---
output: html_document
editor_options: 
  chunk_output_type: console
---
Analysis of pilot data
analysisID: agan0005

For each of the 3 datatsets (protein, metabolite, species) find out which vairables (molecule or species) are changing most between the groups.

This will give a 3 lists:
(1) discriminating proteins; (2) discriminating metabolites; (3) discriminating species.

We will then do combinations of these in the 06 code.

```{r}

manifestFile <- "../input/metadata.2groups.xlsx"

```

Load library to begin analysis

```{r}

#install.packages("pROC")

library("readxl")
library("dplyr")
library("pROC")

```

Define a function to perform one-way ANOVA over columns of a dataframe

```{r}

run_glm <- function (infile, manifestFile) {
  
  # read-in sample manifest
  sampleKey <- read_excel(manifestFile)
  colnames(sampleKey) <- c("sampleName", "sampleGroup")
  
  # read-in metabolite file
  inputData <- read.table(infile, header = TRUE, check.names = F, sep="\t")

  if (length(inputData[,1]) != 0) {
    # remove the unwanted columns
    row.names(inputData) <- inputData[,1]
    inputData[,1:2] <- NULL
  
    # transpose
    inputData2 <- as.data.frame(t(inputData))
  
    # create data.frame of the type:
    # sampleGroup <T> moleculeValue <END>
  
    inputData3 <- merge(sampleKey, inputData2, by.x="sampleName", by.y="row.names")
    inputData3$sampleGroup <- as.factor(inputData3$sampleGroup)
  
    # run linear model
  
    returnValues <- data.frame(ind=character(), glm.pvalue=double(), roc.auc=double())
  
    for (molecule in 3:length(colnames(inputData3))) {
    
      inputData4 <- inputData3[,c(2, molecule)]
      colnames(inputData4)[2] <- "value"
    
      glmResults <- glm(formula = sampleGroup ~ value,
                     data = inputData4,
                     family = binomial)
    
      ## get the pvalue
    p <- coef(summary(glmResults))[,4][2]
    
      ## get the auc
      temp <- cbind(predict(glmResults), inputData4$sampleGroup)
      colnames(temp) <- c("pred", "status")
     dfTemp <- as.data.frame(temp)
      AUC <- auc(roc(dfTemp$status, dfTemp$pred))[1]
    
      returnValues[nrow(returnValues) + 1,] <- c(colnames(inputData3)[molecule], p, AUC)
    }

    outputData <- merge(returnValues, inputData, by.x="ind", by.y="row.names")
  } else {
    
    returnValues <- data.frame(ind=character(), glm.pvalue=double(), roc.auc=double())
    outputData <- inputData
  }
  return (outputData)
}

```


Metabolite glm

```{r}

## loop over the pvalues

for (pvalue in c(0.01, 0.05)) {
  
  metaboliteFile <- paste0("../output/norm.filtered/metabolite.norm.filtered.", pvalue, ".txt")
  
  # get pvalue and AUC from glm and roc
  outputData <- run_glm(metaboliteFile, manifestFile)

  # write to output
  write.table(outputData, file=paste0("../output/univariable-modelling/metabolite.unimodel.all.", pvalue, ".txt"), row.names=F, col.names=T, sep="\t")

  # write filtered to output
  for (cutoff in c("0.05")) {
  
    filteredOutputData <- dplyr::filter(outputData, glm.pvalue < cutoff)
    write.table(filteredOutputData, file=paste0("../output/univariable-modelling/metabolite.unimodel.all.", pvalue, ".", cutoff, ".txt"), row.names=F, col.names=T, sep="\t")
  
  }
}


```

Protein glm

```{r}

## loop over the pvalues

for (pvalue in c(0.01, 0.05)) {
  
  proteinFile <- paste0("../output/norm.filtered/protein.norm.filtered.", pvalue, ".txt")
    
  # get pvalue and AUC from glm and roc
  outputData <- run_glm(proteinFile, manifestFile)

  # write to output
  write.table(outputData, file=paste0("../output/univariable-modelling/protein.unimodel.all.", pvalue, ".txt"), row.names=F, col.names=T, sep="\t")

  # write filtered to output
  for (cutoff in c("0.05")) {
  
    filteredOutputData <- dplyr::filter(outputData, glm.pvalue < cutoff)
    write.table(filteredOutputData, file=paste0("../output/univariable-modelling/protein.unimodel.all.", pvalue, ".", cutoff, ".txt"), row.names=F, col.names=T, sep="\t")
  
  }
}

```

Species glm

```{r}

for (pvalue in c(0.01, 0.05)) {
  
  speciesFile <- paste0("../output/norm.filtered/species.norm.filtered.", pvalue, ".txt")
    
  # get pvalue and AUC from glm and roc
  outputData <- run_glm(speciesFile, manifestFile)

  # write to output
  write.table(outputData, file=paste0("../output/univariable-modelling/species.unimodel.all.", pvalue, ".txt"), row.names=F, col.names=T, sep="\t")

  # write filtered to output
  for (cutoff in c("0.05")) {

    if (length(outputData[,1])!= 0) {
      
      filteredOutputData <- dplyr::filter(outputData, glm.pvalue < cutoff)
    } else {
      filteredOutputData <- outputData
    }
    write.table(filteredOutputData, file=paste0("../output/univariable-modelling/species.unimodel.all.", pvalue, ".", cutoff, ".txt"), row.names=F, col.names=T, sep="\t")
  
  }
}

```

