---
title: "deseq2, unpaired all norm analysis"
author: "Alex Gibbs"
date: "15/08/2023"
output: html_document
editor_options: 
  chunk_output_type: console
---

Notes:

Double check the targets file on line 26 and line 173, is the right targets file selected?


Set working directory and define the required comparisons

```{r}

setwd("/Users/alexgibbs/OneDrive - Cardiff University/crest/analyses/agan0001/Deseq2/bin/")

#countInputFiles <- c("all.markdup.genecount", "all.rmdup.genecount")
countInputFiles <- c("all.markdup.genecount")

fpkmCountfilters <- c("nofilter", 0, 0.5, 5)
#fpkmCountfilters <- c(5)
##############################################################################################
targets <- read.table("../resources/targets-v2.csv", sep=",", header=T)
##############################################################################################
myComparisons <- c(#"CER_384_LPGP_vs_CER_384_LNGP",
                   #"CER_526_LPGP_vs_CER_526_LNGP",
                   #"CER_625_LPGP_vs_CER_625_LNGP",
                   #"CER_626_LPGP_vs_CER_626_LNGP",
                   #"E6_645_LPGP_vs_E6_645_LNGP",
                  #"E6_661_LPGP_vs_E6_661_LNGP",
                   #"WT_421_LPGP_vs_WT_421_LNGP",
                  #"WT_422_LPGP_vs_WT_422_LNGP",
                  #"WT_609_LPGP_vs_WT_609_LNGP",
                  #"WT_621_LPGP_vs_WT_621_LNGP",
                   #"WT_622_LPGP_vs_WT_622_LNGP",
                   #"E6_LPGP_vs_E6_LNGP",
                   #"E6_LPGP_vs_E6_LNGP",
                  # "WT_LPGP_vs_WT_LNGP",
                   #"E6_LPGP_vs_WT_LPGP",
                   #"E6_LNGP_vs_WT_LPGP"
                   #"E6_LPGP_vs_WT_LPGP",
                  # "CER_LNGP_vs_WT_LNGP"
                   #"E6_LNGP_vs_WT_LNGP"
                   "E6_LPGP_vs_E6_LNGP",
                   "E6_LPGP_vs_WT_LNGP",
                   "E6_LPGP_vs_WT_LPGP",
                   "E6_LNGP_vs_WT_LPGP",
                   "E6_LNGP_vs_WT_LNGP",
                   "WT_LPGP_vs_WT_LNGP")

apadjCutoffs <- c("allgenes", "top50genes", "sig0-01", "sig0-05")

#run without removing any outliers first
#outliers <- NULL

#if there are outliers you wish to remove, input the analysisIDs below
outliers <- c("E6_645_LPGP", "E6_645_LNGP", "CER_625_LPGP", "CER_625_LNGP", "WT_621_LNGP", "WT_621_LPGP", "CER_384_LNGP", "E6_661_LPGP","WT_422_LPGP")
             #"WT_622_LNGP", "CER_626_LNGP", "CER_626_LPGP")
              #"WT_421_LPGP", "WT_422_LPGP", "WT_609_LPGP", "WT_621_LPGP", "WT_622_LPGP", "CER_626_LNGP", "CER_626_LPGP")

################################################################################################
pcaWindows <- c("sampleGroup")#, "day", "pathogen") 
################################################################################################
```

Load relevant libraries

```{r}

library(DESeq2)     # differential analysis

library(ape)        # clustering QC
library(ggplot2)    # PCA plotting
library(ggrepel)

library(dplyr)            # data manipulations

```

Get all genes length from the all.markdup.genecount.txt file

```{r}

rawData <- read.table("../input/all.markdup.genecount.txt", sep="\t", header=T)

geneLengths <- dplyr::select(rawData, ensemblGeneID, geneLength)

```

Define functions used in the code below

```{r}

RunDeseq2 <- function(rc, gl) {
      
    ## create the experimental design object for DeSeq2
    
    exptDesign <- data.frame(
            row.names = colnames(rc),
            condition = sampleGroups)
    
    ## construct experiment object
    
    exptObject <- DESeqDataSetFromMatrix(
          countData = rc,
          colData = exptDesign,
          design = ~ condition)

    ## run the anlysis
        
    analysisObject = DESeq(exptObject)

    ## add gene lengths to the analysis object for the FPKM calculation
    
    geneLengthsTemp <- gl[gl$ensemblGeneID %in% rownames(analysisObject),]
    mcols(analysisObject)$basepairs <- geneLengthsTemp[match(rownames(analysisObject), geneLengthsTemp$ensemblGeneID),]$geneLength
    
    return(analysisObject)    
}


GetUniqueGenes <- function(anr, ty) {
  
  sgenes <- NULL
  anr <- anr[order(anr$pvalue), ]  
  
  if (ty == "allgenes") {
     sgenes <- rownames(anr)
  } else if (grepl("^sig", ty)) {
    ty <- gsub("-", ".", gsub("sig", "", ty))
    
    ## need to get genes that are also NA but significant
    
    for (i in 1:length(anr[,1])) {
    
      if (is.na(anr[i,]$padj)) {
        sgenes <- c(sgenes, rownames((anr[i,]))) 
      } else if ((as.numeric(anr[i,]$padj) < as.numeric(ty))) {
        sgenes <- c(sgenes, rownames((anr[i,]))) 
      } else {
        break
      }
    }
  } else if (grepl("^top", ty)) {
    ty <- gsub("top", "", gsub("genes", "", ty))
    sgenes <- rownames(anr[1:as.numeric(ty),])
  } else {
    stop("cuttoff not recognised")
  }
  
  return (sgenes)
}


    
```


Perform the differential analysis over the count filters

```{r}

## loop over the 2 files under input/

for (countfilter in fpkmCountfilters) {
  
  for (file in countInputFiles) {
  ###############################  ###############################  ###############################  ###############################  ###############################  ###############################
    # open targets file for reading
    targets <- read.table("../resources/targets-v2.csv", sep=",", header=T)
     ###############################  ###############################  ###############################  ###############################  ###############################  ###############################
    ## read input file
    
    rawCounts = read.table(paste("../input/", file, ".txt", sep=""), sep="\t", header=TRUE, check.names=F)
    
    ## remove outliers
    
    if (length(outliers) > 0) {
      
      targets <- targets[!targets$analysisID %in% outliers, ]
      rawCounts <- rawCounts[, !(names(rawCounts) %in% outliers)]
    }
    
    ## remove any minus sign from the column names
    
    colnames(rawCounts) <- sub("-", ".", colnames(rawCounts))
    
    # parse out the 3 first columns of annotation
    
    annotationNames <- rawCounts[,c(1:4)]
    
    # get the sample groups and mouse definitions from targets
    ################################################################################################################################
    sampleGroups <- targets$sampleGroup
    ################################################################################################################################
    ## RUN ANALYSIS 
    
    # add rownames to the raw data, and remove all annotation columns
    
    rownames(rawCounts) = rawCounts[,1]
    rawCounts[,c(1:4)] <- NULL
    
    ## check the order of the targets files and input counts file are identical
    
    if (!identical(as.character(targets$analysisID), colnames(rawCounts))) {
      rawCounts <- dplyr::select(rawCounts, as.character(targets$analysisID))
    }
    
    ## run analysis without filters
    
    analysisObject <- RunDeseq2(rawCounts, geneLengths)
      
    ## get FPKM normalised data from the analysis object
    
    rawCounts <- counts(analysisObject, normalized = FALSE)
    fpkmNormalisedCounts <- fpkm(analysisObject, robust = TRUE)
    
    if (!identical(rownames(rawCounts), rownames(fpkmNormalisedCounts))) {
      stop()
    }
    
    ## filter the object by FPKMs
    
    if (countfilter != "nofilter") {
      
      ## exclude any gene where it's value is not greater than "countfilter" over any condition
    
      rawCountsToKeep <- rawCounts[rowSums(fpkmNormalisedCounts > countfilter) >= 1, ]
    
      ## run analysis with filters
    
      analysisObject <- RunDeseq2(rawCountsToKeep, geneLengths)
    }
    
    ## get normalised data from the analysis object
    
    rawCounts <- counts(analysisObject, normalized = FALSE)
    normalisedCounts <- counts(analysisObject, normalized = TRUE)
    fpkmNormalisedCounts <- fpkm(analysisObject, robust = TRUE)
    
    if (!identical(rownames(rawCounts), rownames(normalisedCounts))) { stop() }
    if (!identical(rownames(fpkmNormalisedCounts), rownames(rawCounts))) { stop() }
    

    
    
    ## merge raw and normalised data
    ## add the words "raw.counts" to each column names
    
    colnames(rawCounts) = gsub("^", "raw.counts.", colnames(rawCounts))
    colnames(normalisedCounts) = gsub("^", "norm.counts.", colnames(normalisedCounts))
    colnames(fpkmNormalisedCounts) = gsub("^", "fpkm.norm.counts.", colnames(fpkmNormalisedCounts))
    
     ## print counts
    
   
    annotatedRawCounts <- merge(annotationNames, rawCounts, by.x="ensemblGeneID", by.y="row.names")
    annotatedNormCounts <- merge(annotationNames, normalisedCounts, by.x="ensemblGeneID", by.y="row.names")
    annotatedFPKMCounts <- merge(annotationNames, fpkmNormalisedCounts, by.x="ensemblGeneID", by.y="row.names")
      
    write.table(annotatedRawCounts, paste("../output/", file, ".FPKM", countfilter, ".rawcounts.txt", sep=""), sep="\t", row.names = F, quote = F)
    write.table(annotatedNormCounts, paste("../output/", file, ".FPKM", countfilter, ".normcounts.txt", sep=""), sep="\t", row.names = F, quote = F)
    write.table(annotatedFPKMCounts, paste("../output/", file, ".FPKM", countfilter, ".fpkmnormcounts.txt", sep=""), sep="\t", row.names = F, quote = F)
    
    
    
    
    tempData = merge(rawCounts, normalisedCounts, by="row.names", all=T)
    rownames(tempData) <- tempData[,1]
    tempData[,1] <- NULL
  
    finalData = merge(tempData, fpkmNormalisedCounts, by="row.names", all=T)
    rownames(finalData) <- finalData[,1]
    finalData[,1] <- NULL
  
    printNormCounts <- data.frame(ensembleID=rownames(fpkmNormalisedCounts), fpkmNormalisedCounts)
    ################################################################################################################################################################
    ## circle over the comparisons of everything versus everything
    ################################################################################################################################################################
    for (myComparison in myComparisons) {
      
        numerator <- unlist(strsplit(myComparison, "_vs_"))[1]
        denominator <- unlist(strsplit(myComparison, "_vs_"))[2]
        
        print(myComparison)
                            
        result <- as.data.frame(results(analysisObject, contrast=c("condition", numerator, denominator), independentFiltering=TRUE, pAdjustMethod="BH"))
        
        slimData <- as.data.frame(cbind(rownames(finalData), finalData))
                       
        colnames(slimData)[1] <-  "tracking_id"
             
         ## get only the genes of interest
        
        slimData <- dplyr::select(slimData, tracking_id, colnames(dplyr::select(slimData, paste("raw.counts", unlist(subset(targets, sampleGroup == numerator, analysisID)), sep="."), paste("raw.counts", unlist(subset(targets, sampleGroup == denominator, analysisID)), sep="."), paste("norm.counts", unlist(subset(targets, sampleGroup == numerator, analysisID)), sep="."), paste("norm.counts", unlist(subset(targets, sampleGroup == denominator, analysisID)), sep="."), paste("fpkm.norm.counts", unlist(subset(targets, sampleGroup == numerator, analysisID)), sep="."), paste("fpkm.norm.counts", unlist(subset(targets, sampleGroup == denominator, analysisID)), sep="."))))
                          
    
        printData <- merge(slimData, result, by.x = "tracking_id", by.y = "row.names", all = T)

        ## add annotation
            
        printData <- merge(annotationNames, printData, by.x="ensemblGeneID", by.y="tracking_id") 
                
        printData <- printData[order(printData$pvalue), ]    
        
        ## add second header
        
        printData %>% mutate_if(is.factor, as.character) -> printData2
        printData2 <- rbind(colnames(printData), printData2)
        
        ## get group names for final output file
    
        temp <- colnames(slimData[-1])
        temp <- gsub("^raw.counts.", "", gsub("^norm.counts.", "", gsub("^fpkm.norm.counts.", "", temp)))
        
        outputColHeadings <- c(colnames(annotationNames), as.vector(targets$sampleGroup[match(temp, targets$analysisID)]), colnames(result))
  
        printData2 <- rbind(outputColHeadings, printData2)
        
        ## comment on outliers removed
        commentLine <- rep("", length(outputColHeadings))
        commentLine[1] <- paste("## Outliers removed: ", paste(outliers, collapse=" "), sep="")
        printData2 <- rbind(commentLine, printData2)
        
        write.table(printData2, file=paste("../output/", numerator, "_vs_", denominator, ".", file, ".FPKM", countfilter, ".allnorm.txt", sep=""), col.names=F, row.names=F, sep="\t", quote=F)
                    
     }
    
    ## perform ANOVA for the PCA: looking for changing genes across the entire experiment
    
    anovaResult <- as.data.frame(results(analysisObject, pAdjustMethod="BH"))
    anovaResult <- anovaResult[order(anovaResult$pvalue), ]  
    
    ## plot all genes PCA
    
    for (dirName in apadjCutoffs) {
      
      print(dirName )
      
      ## get unique genes
      uniqueSignificantGenes <- GetUniqueGenes(anovaResult, dirName);
      
      if (length(uniqueSignificantGenes) > 5) {
        
        ## create a directory under output if not there already
      
        if (!file.exists(paste("../output/", dirName, sep=""))) {
          dir.create(file.path("../output/", dirName))
        }
        
        
        ## create subset of significant genes
        
        pcaDD <- subset(fpkmNormalisedCounts, rownames(rawCounts) %in% uniqueSignificantGenes)
        
        
        ## create object for hierarchical clustering
        
        hierData <- pcaDD
    
        ## change label name to feature sampleGroup
        b <- NULL
    
        for (a in gsub("^fpkm.norm.counts.", "", colnames(hierData))) {
          b <- c(b, paste(as.vector(targets$sampleGroup[targets$analysisID == a]), a, sep=" "))
        }
    
        colnames(hierData) <- b
    
        d <- cor(hierData, method="pearson")
        hc <- hclust(dist(1-d))
    
        ## write to outputfile
    
        hierData <- data.frame(ensembleGeneID=rownames(hierData), hierData)
        
        aa <- subset(rawData, ensemblGeneID %in% rownames(hierData), select=c(ensemblGeneID, geneName))
        bb <- merge(aa, hierData,  by.y="row.names", by.x="ensemblGeneID")
        bb[,3] <- NULL
                   
        write.table(bb, file=paste("../output/", dirName, "/hcluster.", file, ".FPKM", countfilter, ".FPKM.", dirName, ".txt", sep=""), col.names=T, row.names=F, sep="\t", quote=F)
    
    
        ## create pca object
    
        pca <- prcomp(t(pcaDD), center=TRUE, scale=F)
        scores <- data.frame(targets$analysisID, pca$x[,1:2])
    
        # write PCA coordinates to file
    
        write.table(scores, paste("../output/", dirName, "/PCA.", file, ".FPKM", countfilter, ".coord.", dirName, ".txt", sep=""), sep="\t", row.names = F, quote = F)
       
        # write loadings to a file
       
        aload <- abs(pca$rotation)
        loadings <- sweep(aload, 2, colSums(aload), "/")
        loadings <- cbind(row.names(loadings), loadings)
        write.table(loadings, paste("../output/", dirName, "/PCA.", file, ".FPKM", countfilter, ".load.", dirName, ".txt", sep=""), sep="\t", row.names = F, quote = F)
              
        # write percentage contributions to file
       
        write.table(summary(pca)$importance[2,], paste("../output/", dirName, "/PCA.", file, ".FPKM", countfilter, ".perc.", dirName, ".txt", sep=""), sep="\t", row.names = T, quote = F)
       
        ## open the PDF
      
        pointSize <- 0.4
    
        pdf(paste("../output/", dirName, "/PCA.", file, ".FPKM", countfilter, ".", dirName, ".pdf", sep=""), onefile=T)
    
        ## print dendrogram
        
        tiplabs <- as.phylo(hc)$tip.label
        numbertiplabs<-length(tiplabs)
        colourtips<-rep("blue",numbertiplabs)
        colourtips[grep("CTR2",tiplabs)]<-"red"

        plot.phylo(as.phylo(hc), type="p", show.node.label=TRUE, main=paste(file, "plotting FPKMs using [", length(uniqueSignificantGenes), "] genes.  Outliers removed are: [" , paste(outliers, collapse = " "), "]", sep=" "), cex=pointSize, cex.main=pointSize, tip.color=colourtips)
        
        ## print PCA windows
        
        for (x in pcaWindows) {
          print(qplot(x=PC1, y=PC2, main = paste(file, "plotting FPKMs using [", length(uniqueSignificantGenes), "] genes.  Outliers removed are: [" , paste(outliers, collapse = " "), "]", sep=" "), data=scores, label=factor(targets$analysisID), colour=factor(targets[[x]])) + geom_text_repel(size=2.5) +  scale_colour_discrete(name=x) + theme(plot.title = element_text(size = 5))  )
        }
    
        dev.off()
      }
    }
  }
}

```

Remake PCA plots with different labels etc
```{r}
 
        for (x in pcaWindows) {
          print(qplot(x=PC1, y=PC2, main = paste(file, "plotting FPKMs using [", length(uniqueSignificantGenes), "] genes.  Outliers removed are: [" , paste(outliers, collapse = " "), "]", sep=" "), data=scores, label=factor(targets$sampleGroup), colour=factor(targets[["sampleGroup"]])) + geom_text_repel(size=2.5) +  scale_colour_discrete(name=x) + theme(plot.title = element_text(size = 5))  )
        }

```
